# DESeq analysis of mRNA-seq data for PCBC

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(DESeq2)
library(dplyr)
library(plyr)
library(reshape2)
library(synapseClient)
library(knitr)
library(ggplot2)
library("BiocParallel")
register(MulticoreParam(8))

source("./libs.R")
synapseLogin()

opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=8)

htseqfiledir <- "../snakemake/output"

# Save myself to synapse
fn <- File("./DESeq2_mRNA.Rmd", parentId="syn2838880")
myself <- synStore(fn)

```

Read in the metadata from annotations on htseq-counts data:

```{r readsamplemetadata}
sampleTableQuery <- "select UID,name,id,C4_Cell_Line_ID,High_Confidence_Donor_ID,Originating_Lab,Gender,Diffname_short,Reprogramming_Gene_Combination,Reprogramming_Vector_Type,Cell_Line_Type,Cell_Type_of_Origin from file where fileType=='count' AND dataType=='mRNA'"

sampleTable <- synapseQuery(sampleTableQuery)
sampleTable <- tbl_df(sampleTable)

colnames(sampleTable) <- gsub("file.", "", colnames(sampleTable))

# Update a Donor ID column for normalization purposes
sampleTable <- mutate(sampleTable,
                      Donor_ID=factor(ifelse(High_Confidence_Donor_ID == "", 
                                      C4_Cell_Line_ID, 
                                      High_Confidence_Donor_ID)),
                      C4_Cell_Line_ID=factor(C4_Cell_Line_ID),
                      Originating_Lab=factor(Originating_Lab),
                      Gender=factor(Gender),
                      Diffname_short=factor(Diffname_short),
                      Reprogramming_Gene_Combination=factor(Reprogramming_Gene_Combination),
                      Reprogramming_Vector_Type=factor(Reprogramming_Vector_Type),
                      Cell_Type_of_Origin=factor(Cell_Type_of_Origin),
                      Cell_Line_Type=factor(Cell_Line_Type))

# Reorder the columns for DESeq reading
sampleTable <- sampleTable[, c("UID", "name", "C4_Cell_Line_ID", "Donor_ID" , "Gender", "Originating_Lab", 
                               "Diffname_short", "Reprogramming_Gene_Combination", "Reprogramming_Vector_Type",
                               "Cell_Line_Type", "Cell_Type_of_Origin")]

# Covariates to consider
covariates <- c("Gender", "Originating_Lab", "Cell_Type_of_Origin", "Cell_Line_Type",
                "Reprogramming_Gene_Combination", "Reprogramming_Vector_Type")

```

The sample table is filtered by the mutually agreed upon list from 2014-11-25.

```{r filtersamples, echo=TRUE, results='asis'}
uniqueDonorSCFile <- synGet("syn2875170")
uniqueDonorList <- read.delim(uniqueDonorSCFile@filePath, col.names="UID")
sampleTable <- filter(sampleTable, UID %in% uniqueDonorList$UID)
sampleTable <- droplevels(sampleTable)
kable(sampleTable)
```

### DESeq analysis performed for each factor of interest

The covariates under consideration are:

`r paste(covariates, collapse=",")`

#### Perform likelihood ratio tests for whole covariate

First I compare a model (e.g., `count ~ Gender`) to a null model (e.g., `count ~ 1`) to assess an overall effect for each covaraite.

```{r lrtmodels}
allResLRT <- mlply(covariates, dodeseqLRT, sampleTable=as.data.frame(sampleTable), htseqfiledir=htseqfiledir)
names(allResLRT) <- covariates
```

```{r nsigdf}
alpha <- 0.1
nsigdf <- ldply(allResLRT, 
                function(x) data.frame(numSigGenes=nrow(subset(x, padj < alpha))),
                .id="covariate")
```

Number of significantly differentially expressed genes for overall factor comparisons (likelihood ratio test between full model and reduced model, q < `r alpha`):

```{r nsigtbl, echo=FALSE, results='asis'}
nsigdf$X1 <- covariates
colnames(nsigdf) <- c("covariate", "numSigGenes")
kable(nsigdf)
```

```{r extractresultsLRT}
finalResLRT <- ldply(allResLRT, function(y) transform(data.frame(y), 
                                                      GeneSymbol=rownames(as.data.frame(y))),
                     .id="LRT")
finalResLRT$X1 <- rep(covariates, sapply(allResLRT, nrow))
colnames(finalResLRT)[1] <- "LRT"
```

#### Perform pairwise tests within each covariate

Enumerate all pairwise tests within each covariate (e.g., within `Reprogramming_Gene_Combination` we may compare `OSKM` and `OSKM-NLT`). Only groups within each covariate with at least 3 samples are considered.

```{r pairwisetests}
allRes <- mlply(covariates, dodeseqPairwise, sampleTable=as.data.frame(sampleTable), htseqfiledir="../snakemake/output")
names(allRes) <- covariates
```

Extract all the results from all pairwise comparisons and build a table.

```{r extractresults}
finalRes <- llply(allRes, function (x) ldply(x, function(y) mutate(data.frame(y), 
                                                                   GeneSymbol=rownames(as.data.frame(y))),
                                             .id="pairwise"))
finalRes <- ldply(finalRes, .id="covariate")
```

As a conservative adjustment, I also concatenate all test statistics (nominal p-values) and adjust them globally.

```{r globalpadj}
finalRes <- mutate(finalRes,
                   pglobaladj=p.adjust(pvalue, method="BH"))
```

## Save data to Synapse

```{r lrtpvals, message=FALSE, warning=FALSE, echo=FALSE}
pvalueTableLRT <- dcast(finalResLRT, GeneSymbol ~ LRT, value.var="padj")
write.csv(pvalueTableLRT, "./DESeq_pvalueTableLRT.csv", row.names=FALSE, quote=FALSE)

fn <- File("./DESeq_pvalueTableLRT.csv", parentId="syn2838880")
fn <- synStore(fn, executed=myself@properties$id, used="syn2822494")
```

A table of likelihood ratio test adjusted p-values is at `r fn@properties$id`.

```{r pvalglobaladjtbl, message=FALSE, warning=FALSE, echo=FALSE}
pvalueTableGlobalAdj <- dcast(finalRes, GeneSymbol ~ pairwise, value.var="pglobaladj")
write.csv(pvalueTableGlobalAdj, "./DESeq_pvalueTableGlobalAdj.csv", row.names=FALSE, quote=FALSE)

fn <- File("./DESeq_pvalueTableGlobalAdj.csv", parentId="syn2838880")
fn <- synStore(fn, executed=myself@properties$id, used="syn2822494")
```

A table of global adjusted p-values is at `r fn@properties$id`.

```{r pvaladjtbl, message=FALSE, warning=FALSE, echo=FALSE}
pvalueTableAdj <- dcast(finalRes, GeneSymbol ~ pairwise, value.var="padj")
write.csv(pvalueTableAdj, "./DESeq_pvalueTableAdj.csv", row.names=FALSE, quote=FALSE)

fn <- File("./DESeq_pvalueTableAdj.csv", parentId="syn2838880")
fn <- synStore(fn, executed=myself@properties$id, used="syn2822494")
```

A table of pairwise adjusted p-values is at `r fn@properties$id`.

```{r fullrestbl, message=FALSE, warning=FALSE, echo=FALSE}
# write.table(finalRes, "DESeq_finalResultsTable.tsv", sep="\t", row.names=FALSE, quote=TRUE)
# fn <- File("./DESeq_finalResultsTable.tsv", parentId="syn2838880")
# fn <- synStore(fn, executed=myself@properties$id, used="syn2822494")
# # Full results table of the pairwise comparisons, including log fold changes is at `r fn@properties$id`. 
```

For viz purposes, compute regularized log transformation on counts.

```{r rlog}
dds <- DESeqDataSetFromHTSeqCount(as.data.frame(sampleTable),
                                  directory=htseqfiledir,
                                  design=~1)

rld <- rlog(dds, blind=TRUE, fast=FALSE)
```

## Data visualization

DESeq2 provides a novel method for data visualization (regularized log, or `rlog`) which deal with effects of heteroscedasticity endemic in expression assays.

### PCA of rlog-transformed data colored by originating lab using 500 most variable genes
```{r samplepcalab}
plotPCA(rld, intgroup = c("Originating_Lab")) + theme_bw()
```

### PCA of rlog-transformed data colored by Gender using 500 most variable genes
```{r samplepcagender}
plotPCA(rld, intgroup = c("Gender")) + theme_bw() + theme(legend.position="top")
```

### PCA of rlog-transformed data colored by Cell_Line_Type using 500 most variable genes
```{r samplecelllinetype}
plotPCA(rld, intgroup = c("Cell_Line_Type")) + theme_bw() + theme(legend.position="top")
```

### PCA of rlog-transformed data colored by Cell_Type_of_Origin using 500 most variable genes
```{r samplecelltypeoforigin}
plotPCA(rld, intgroup = c("Cell_Type_of_Origin")) + theme_bw() + theme(legend.position="top")
```

### PCA of rlog-transformed data colored by Reprogramming_Vector_Type using 500 most variable genes
```{r samplepcareprogvectortype}
plotPCA(rld, intgroup = c("Reprogramming_Vector_Type")) + theme_bw() + theme(legend.position="top")
```
